---
title: "STAT545_hw05_JasmineLib"
output: github_document
---
cowplot
gridextratidy

# Homework 5: Factor and Figure Management
```{r}
library(gapminder)
library(tidyverse)
library(plotly)
```


### Part 1: Factor Management

Task 1.1: Drop a factor to remove Oceania. Remove unused factor levels. Provide concrete information before and after removing these rows and the number of levels of affected factors.
explore the effects of arrange()
explore effects of reordering a factor and try this coupled with arrange() How does this affect figure output?
```{r}
str(gapminder)
gapminder %>% 
  filter(continent != "Oceania") %>% 
  ggplot(aes(continent)) + geom_bar() + 
  scale_x_discrete(drop = FALSE) #shows "hidden factor Oceania"

?fct_infreq()

gapminder$country %>% 
  filter(continent != "Oceania") %>% 
  droplevels() %>%  #removes oceania.
  mutate(country = fct_infreq(gdpPercap, country)) %>% 
  ggplot(aes(country, gdpPercap)) +
  geom_point() 
```

### Part 2: File I/O

```{r}
oil_consumption = read_csv(file ="Oil_Consumption_per_capita.csv")

head(oil_consumption)

?saveRDS()


oil_consumption_2007 = oil_consumption %>% 
  select(`Oil Consumption per capita (tonnes per year)`, "2007") %>% 
  mutate(country = `Oil Consumption per capita (tonnes per year)`, 
         tonnes_per_capita_2007 = `2007`) %>% 
  select(country, tonnes_per_capita_2007)

head(oil_consumption_2007)

```

```{r}
gapminder_2007 = gapminder %>% 
  filter (year == "2007") %>% 
  select(country = "country", pop, gdpPercap, continent) #select the population and gdpPercap data for the year 2007. 
  
gapminder_2007 %>% 
  head() %>% 
  knitr::kable()
```

```{r}
gapminder_oil_2007 = left_join(gapminder_2007, oil_consumption_2007, by= "country") %>% 
  na.omit() 

gapminder_oil_2007
```


### Part 3: Visualization Design
 






### More? 