---
title: "STAT545_hw05_JasmineLib"
always_allow_html: yes
output: github_document

---
cowplot
gridextratidy

# Homework 5: Factor and Figure Management
```{r}
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(gridExtra))

```


### Part 1: Factor Management

####Task 1.1: Drop a factor 
- remove Oceania. 
- Remove unused factor levels. 
- Provide concrete information before and after removing these rows and the number of levels of affected factors.
```{r Drop Gapminder}
levels(gapminder$continent)
nlevels(gapminder$country)

gapminder_no_oceania = gapminder %>% 
  filter(continent != "Oceania") %>% 
  droplevels()

levels(gapminder_no_oceania$continent)
nlevels(gapminder_no_oceania$country)

```

  
Conclusion:  

| Function Applied  | Resulting Continent Factor |  Resulting Country Factor |
| ------------- | ------------- |----------|
| Filter out Oceania  | 5 continents  |   142 countries     |
| Drop Levels  | 4 continents |   140 countries     |



####Explore the effects of arrange( ).   
-  Does merely arranging the data have any effect on, say, a figure?
```{r Arrange Effects}
#using arrange in a table:
gapminder %>% 
  filter(continent == "Americas" & year == 2007) %>% 
  arrange(desc(lifeExp)) %>% 
  head() %>% 
  knitr::kable()

#before arrange:
gapminder %>% 
  filter(continent == "Americas" & year == 2007) %>% 
  ggplot(aes(lifeExp, country)) + 
  geom_point() +
  ggtitle("Life Expectancy of Countries in Americas in 2007") +
  theme_bw() +
  labs(x = "Life Expectancy (Years)", y = "Country")

#using arrange(desc())
gapminder %>% 
  filter(continent == "Americas" & year == 2007) %>% 
  arrange(desc(lifeExp)) %>% 
  ggplot(aes(lifeExp, country)) + 
  geom_point() +
  ggtitle("Life Expectancy of Countries in Americas in 2007") +
  theme_bw() +
  labs(x = "Life Expectancy (Years)", y = "Country")

```

Conclude:   
- using arrange(desc()) works when making tables  
- both figures are the same, despite one of them having an arrange(desc()) function   
- I conclude that the arrange function has no effect on figures. 



####Explore effects of reordering a factor   
- try this coupled with arrange()   
- How does this affect figure output?
```{r Reorder Factor}
gap_america_2007 = gapminder %>% 
  filter(continent == "Americas" & year == 2007)

#fct_reorder first, then arrange:

gap_america_2007 %>% 
  mutate(country = fct_reorder(country, lifeExp)) %>% 
  arrange(pop) %>% 
  ggplot(aes(lifeExp, country)) + 
  geom_point(color = "dodgerblue3") +
  theme_bw() +
  ggtitle("Life Expectancy in the Americas in 2007") +
  labs(x = "Life Expectancy (Years)", 
       y = "Country")

#arrange first, then fct_reorder:

gap_america_2007 %>% 
  arrange(pop) %>% 
  mutate(country = fct_reorder(country, lifeExp)) %>% 
  ggplot(aes(lifeExp, country)) + 
  geom_point() +
  geom_point(color = "dodgerblue3") +
  theme_bw() +
  ggtitle("Life Expectancy in the Americas in 2007") +
  labs(x = "Life Expectancy (Years)", 
       y = "Country")

```
Conclude:    
- Succesfully sorted by life expectancy
- The fct_reorder function sorted by order of life expectancy, while arrange tried to sort by population size.  
- when using both the arrange() function and the fct_reorder() function, the fct_reorder function takes precedence  
- This was the case regardless of the order in which the two functions are called (both graphs are the same)
- Canada has the highest life expectancy out of all the Americas, while Haiti has the lowest.


### Part 2: File I/O  
- Explore read_csv( ):   
- Used data downloaded from gapminder.org/data to read in a file   
- upload data here

```{r}
oil_consumption = read_csv(file ="Oil_Consumption_per_capita.csv")

oil_consumption_2007 = oil_consumption %>% 
  select(`Oil Consumption per capita (tonnes per year)`, "2007") %>% 
  mutate(country = `Oil Consumption per capita (tonnes per year)`, 
         tonnes_per_capita_2007 = `2007`) %>% 
  select(country, tonnes_per_capita_2007)

#check to ensure csv file was properly loaded:
head(oil_consumption_2007) %>% 
  knitr::kable(col.names = c("Country ", "Tonnes per Capita") )

```

- Next, make a corresponding gapminder subset which we will use to join:
```{r}
gapminder_2007 = gapminder %>% 
  filter (year == "2007") %>% 
  select(country = "country", pop, gdpPercap, continent, lifeExp) 
```
  
- Perform a join:
```{r}
gapminder_oil_2007 = left_join(gapminder_2007, oil_consumption_2007, by= "country") %>% 
  na.omit() 

str(gapminder_oil_2007$continent)
str(gapminder_oil_2007$country)
nlevels(gapminder$country)

```
 
Conclude:   
- after the join, we have 5 levels of continents, as expected  
- after the join, the country factor was changed to a character variable. I will address this later on in the assignment.   
- the number of levels is still 142, suggesting that oil use data was complete and available for all countries in gapminder.

####Further Testing of I/O with help from [Class Notes](http://stat545.com/block026_file-out-in.html#retaining-factor-levels-upon-re-import)

```{r}
#first make the country variable into a factor (in order to look at how factors are affected in I/O)
gapminder_oil_2007$country = as.factor(gapminder_oil_2007$country)
#now we check that this is indeed a factor.
str(gapminder_oil_2007$country) 

test_export = gapminder_oil_2007 %>% 
  mutate(country = fct_reorder(country, tonnes_per_capita_2007))

#make a tibble with original country level
check_order_of_countries = tibble(before_export = head(levels(test_export$country)))
                                  
#write to file using write_csv, saveRDS, and dput: 
write_csv(test_export, "test_export.csv")
saveRDS(test_export, "test_export.rds")
dput(test_export, "test_export_dput.txt")

#import corresponding files using read_csv, readRDS, and dget: 
test_import_csv = read_csv("test_export.csv") %>% 
  mutate(country = factor(country))
test_import_rds = readRDS("test_export.rds")
test_import_dget = dget("test_export_dput.txt")

#look at the first few countries listed to see if the non-alphabetical order "survived" the round trip of writing then reading back in:
check_order_of_countries = check_order_of_countries %>% 
  mutate(imported_by_csv = head(levels(test_import_csv$country)), 
         imported_by_RDS = head(levels(test_import_rds$country)),
         imported_by_dput_dget = head(levels(test_import_dget$country)))
  
check_order_of_countries %>% 
  knitr::kable(col.names = c("Before Export/Import ", "by CSV", "by RDS", "by dput dget"))

```
  
Conclude:  
- using read csv and write csv will NOT preserve changes in factor order  
- using read RDS and write RDS will preserve changes in factor order  
- using dput and dget will preserve changes in factor order.

###Part 3: Visualization Design  
- Here is an old graph from an earlier assignment:
```{r}
 ggplot(gapminder_2007, aes(continent, gdpPercap)) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  geom_violin(fill = "darkslategray1", alpha = 0.5) +
  geom_jitter(alpha = 0.33) +
  xlab("Continent")+
  ylab("GDP Per Capita") +
  ggtitle("GDP Per Capita by Continent in 2007")
```

- In this old graph it's hard to tell visually which continent has a higher mean GDP
- I was not able to tell the difference between America vs. Asia, and Europe vs. Oceania. 
- notice the y-scales are not very easy to read

Using new knowledge, I can make various types of new plots:   

```{r}

gdpPercap_by_continent_updated = gapminder_2007%>% 
  mutate(continent = fct_reorder(gapminder_2007$continent, gdpPercap)) %>% 
 ggplot( aes(continent, gdpPercap)) +
  scale_y_log10(labels = dollar_format()) +
  geom_violin(fill="wheat", color="slateblue4",alpha = 0.5) +
  geom_jitter(alpha = 0.33) +
  xlab("Continent")+
  ylab("GDP Per Capita") +
  ggtitle("GDP Per Capita by Continent in 2007") +
  theme_bw() 

gdpPercap_by_continent_updated
```
  
Changes:  
- nicer colours    
- updated theme  
- y-axis easier to read   
- Mean GDP per capita for each continent in ascending order


###Part 3: Visualization Design Continued 

Other plots that we have learned to make: 
```{r}

plot_oilvsgdp = gapminder_oil_2007 %>% 
  ggplot(aes(gdpPercap, tonnes_per_capita_2007)) + 
  geom_point(aes(colour = pop)) +
  scale_x_log10(labels = dollar_format())+
  scale_colour_distiller(
    trans="log10",
    breaks = 10^(0:9),
    labels = comma_format(),
    palette = "YlOrRd"
  ) +
  scale_y_continuous(breaks = c(0,2,4,6,8,10))+
  theme_bw()+
  ggtitle("Tonnes of Oil per Capita by GDP per Capita in 2007")+
  labs(y = "Oil Use Per Capita (Tonnes)", x = "GDP Per Capita")

plot_oilvsgdp

#Table showing the 5 countries with the lowest oil usage per capita in 2007
gapminder_oil_2007_min = gapminder_oil_2007%>% 
  select (country,tonnes_per_capita_2007) %>% 
  top_n (-5) %>% 
  arrange(tonnes_per_capita_2007)

#Table showing the 5 countries with highest oil usage per capita in 2007 
gapminder_oil_2007_max = gapminder_oil_2007 %>% 
  select (country, tonnes_per_capita_2007) %>% 
  top_n (5) %>% 
  arrange(desc(tonnes_per_capita_2007) )

max_oil_tbl = tableGrob(gapminder_oil_2007_max,  cols = c("Country", "Tonnes per Capita"))
min_oil_tbl = tableGrob(gapminder_oil_2007_min, cols = c("Country", "Tonnes per Capita"))

#use gridarrange to nicely display countries using the most and least oil per capita.
grid.arrange(top="Worldwide Oil Use 2007", plot_oilvsgdp, arrangeGrob(max_oil_tbl, min_oil_tbl, ncol = 2),
             heights=c(2.5/4,1.5/4),ncol = 1)

```

Conclude:   
- There is a trend showing increasing oil usate as GDP per Capita increases.   
- Using the R Colour Distiller, we also notice that countries with a higher population tend towards lower oil usage    
- Countries with a low, or mid-range population size appear to use more oil.
- a plotly graph would be useful for a user to zoom in and see which countries are where.

###Self-contained plotly code:  
- Refer to file called "plotly_hw05_JasmineLib.Rmd" to run the plotly graph

###Part 4: Writing figures to file  

```{r}
plot_oilvsgdp
ggsave("tonnes_per_capita_vs_gdpPercap.jpeg", scale = 1, width = 6, height = 4, units = "in")

gdpPercap_by_continent_updated
ggsave("gdpPercap_by_continent_updated_plot.jpeg", scale = 1, width = 6, height = 4, units = "in")

```


### Bonus -  Creating a new factor:
```{r}
gapminder_2007_sport = gapminder_2007%>% 
  filter(country == "Brazil" | country == "Canada" |country == "France") %>% 
  mutate(sport = factor(c("Capoeira","Hockey", "Soccer")))
  
levels(gapminder_2007_sport$sport)
str(gapminder_2007_sport)
```

